Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-08-02 paper.tex
corrections and improvements to final version

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. We predict the Virgo, Fornax and \C{M49} clusters to
  be the brightest DM sources and find a particularly low CR induced
  background for Fornax. For a minimum substructure mass given by the
  DM free-streaming scale, the largest DM halo mass maximizes the
  substructure boost over the smoothly distributed annihilation flux:
  for galaxy clusters as the largest collapsed objects, we find a
  boost factor of more than 1000. Since the annihilation flux of
  substructures is mostly contributed by the regions around the virial
  radius, the resulting surface brightness profiles are almost
  flat. This makes it very challenging to detect this flux with
  imaging air Cherenkov telescopes since their sensitivity drops
  linearly with radius and they typically have $5-10$ linear
  resolution elements across a cluster; the gamma-ray satellite \Fermi
  is better suited. Assuming cold dark matter with a substructure mass
  distribution down to an Earth mass and using \Fermi upper limits, we
  rule out the leptophilic models in their present form in 28
  clusters, and limit the boost from Sommerfeld enhancement in the
  most constraining clusters, \C{M49 and Fornax, } to be less than 5. This
  corresponds to a Sommerfeld enhancement in the Milky Way of less
  than 5. Alternatively, if the Sommerfeld enhancement factor is
  realized in Nature, this would imply a limiting substructure mass of
  $\mlim > 10^4\,\msun$ which would be a challenge for structure
  formation. We find it challenging for \Fermi to constrain our
  selection of DM benchmark models in the \C{near future} using clusters
  without improved modelling of extended sources in addition to a
  stacked likelihood analysis. The \Fermi upper limits are, however,
  closing in on our predictions for the CR flux using an analytic
  model based on cosmological hydrodynamical cluster simulations. Thus
  we will soon start to constrain the underlying CR physics such as
  shock acceleration efficiencies or CR transport properties.
\end{abstract}

@@

Among the astrophysical systems which will be very interesting to detect, and
study, with coming gamma-ray detectors (\Fermi, H.E.S.S., MAGIC, VERITAS, and
eventually large detectors like CTA)\footnote{{High Energy Stereoscopic System,
    Major Atmospheric Gamma Imaging Cerenkov Telescope, Very Energetic Radiation
    Imaging Telescope Array System, Cerenkov Telescope Array.}} belong galaxy
clusters. The most promising directions in which to search for a gamma-ray
annihilation signal (from the annihilation process itself, and also the
accompanying bremsstrahlung and inverse Compton (IC) components coming from
charged particles produced in the annihilations) are basically three:

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize already
in Table~\ref{tab:flux_tab} our results for the gamma-ray flux from
four promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo as the prime targets for DM observations, Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.}  For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ $^{(1)}$ & $\B_\rmn{sfe} $$^{(2)}$ &  $\B_\rmn{sub} $$^{(3)}$ \\
         & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ 
         & & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $8.2\times10^{-8}$  & $7.6\times10^{-11}$ & $4.1\times10^{-9}$  
 & \,\,\,\,\,         & $1.2\times10^{-11}$ & $2.8\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $12.9$ & $530/65$  & $1290$ \\
 Perseus \,\,\,\,\,\, & $8.6\times10^{-8}$  & $7.8\times10^{-11}$ & $1.5\times10^{-8}$  
 & \,\,\,\,\,         & $1.2\times10^{-11}$ & $2.9\times10^{-13}$ & $5.5\times10^{-12}$ 
 & $8.6$ & $530/75$  & $1190$ \\
 Virgo                & $1.6\times10^{-6}$  & $1.5\times10^{-9}$ & $1.5\times10^{-8}$  
 & \,\,\,\,\,         & $2.3\times10^{-10}$ & $5.4\times10^{-12}$ & $5.7\times10^{-12}$ 
 & $6.9$  & $530/80$  & $1120$  \\
 Fornax               & $3.5\times10^{-7}$  & $3.2\times10^{-10}$ & $3.1\times10^{-10}$ 
 & \,\,\,\,\,         & $5.1\times10^{-11}$ & $1.2\times10^{-12}$ & $1.1\times10^{-13}$ 
 & $2.4$  & $530/110$  & $890$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The mass of Fornax, Coma, and Perseus are derived from \cite{2007A&A...466..805C}, 
  while the mass of Virgo is derived from \cite{1984ApJ...281...31T}.\\
  (2) The boost due to Sommerfeld enhancement. The first value shows the saturated 
  Sommerfeld boost realized when substructures are present, the latter is the Sommerfeld 
  boost without substructures. See Section II A 1.\\
  (3) We integrate the emission in a cylinder with a radial extent of $2.5
  R_{200}$ along the line-of-sight and an angular size corresponding to $\rvir$
  (both measured from the cluster center).\\
  (4) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  \Fermi (see the following sections for detail).\\
  (5) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (6) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). Furthermore,
since most of gamma-ray flux is expected from come from dense subhalos
at which the velocity dispersion is close to the velocity limit
$v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(\sigv=v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that, $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross-section, but here we have in addition
  included the order unity constraint that make sure that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess at
Earth observed by the PAMELA and \Fermi satellites,
\begin{equation}
%\B_\sfe(\sigv) = 0.266\, \frac{c}{v_\rmn{sat}}\, \frac{v_\rmn{MW}}{\sigv} = 
%121 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.
\C{\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} = 
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.}
\label{eq:B_sfe}
\end{equation}
\C{Here the boost factor needed to explain the PAMELA/\Fermi excess of
  electrons and positrons $\B_\rmn{sfe,MW}=300$, the velocity
  dispersion for the MW $v_\rmn{MW} \approx 220\,$~km/s and the
  velocity dispersion $\sigv$ for clusters is derived in
  \cite{2005RvMP...77..207V}.} Clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution dissipationless DM simulations of MW type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center suffer from dynamical friction
and tidal effects depleting their central number densities. Since the
DM annihilation rate depends on the density squared, the resulting
flux from substructures is boosted compared to the smooth density
distribution. Recent high resolution simulations suggest a flux
enhancement from substructures of more than 100
\cite{2008MNRAS.391.1685S, 2010ApJ...718..899A} for a MW size halo.
We adopt a boost from substructures of 230 which is the enhancement
found in one of the highest resolved MW halos ever simulated. While
there is still a discrepancy in the literature of the exact value of
the predicted boost factor of the DM annihilation luminosity from
substructures in DM halos, this inconsistency starts to become
resolved, apparently converging toward predictions at the high end,
e.g. those by the Aquarius simulations \protect
\cite{2008MNRAS.391.1685S, 2008Natur.456...73S}. The initial
suggestions of a small {\em total} boost ($<R_{200}$) of order unity
by the Via Lactea simulations \cite{2008JPhCS.125a2008K} made
assumptions of computing the boost assuming that the substructure
luminosity follows that of the smooth DM distribution. Physically,
this would imply substructure to follow the smooth DM distribution and
a radially independent concentration parameter of substructures; both
following from neglecting tidal mass loss of satellites. Tidal losses,
however, are at work in simulations and cause the radial number
density to be anti-biased with respect to the host's mass density
profile yielding a significantly flatter subhalo distribution compared
to the smooth DM distribution \cite{2010ApJ...718..899A}. This
increases the substructure boost preferentially in the outer parts of
DM halos with a factor that ranges between 20 and 1000 (for dwarfs and
galaxy clusters), hence depending sensitively on the host halo mass
(see Figure~\ref{fig:radial_lum}). \C{It should be noted that the very
  recent {\it Phoenix} simulations by Chen et
  al. \cite{2011arXiv1107.1916G} confirms the large enhancement due to
  substructures in clusters. In fact they find a boost of 1125 for a
  Coma like cluster which agrees within 10\% with our estimated boost
  (see e.g. Table~\ref{tab:flux_tab}).}

@@

We now derive the squared density profile for the substructures
using
\begin{eqnarray}
\rho_\rmn{sub}^2(r) &=& \frac{\dd L_\rmn{sub}}{\dd V} \frac{1}{A_{\rmn{dm}}}\,,\label{eq:rho_sub}\\
L_{200\sm} &=&  A_{\rmn{dm}} \int_{\rvir} \dd r' 4\pi r'^2 \rho^2(r')\,,\label{eq:Lsm}\\
 \frac{\dd L_\rmn{sub}}{\dd V} &=& a_0\,a_1\,C(\mvir)\,x^{g(x)}\,
\left[\frac{1+a_2\log(x)}{4\pi\,r^3}\right]\,,\nonumber\\
\\
x &=& r/\rvir\,,\quad g(x) = a_2+a_1\,x^{a_2}\,.\label{eq:xvir}
\end{eqnarray} 
Here, $A_{\rmn{dm}} = q_{\rmn{sm},i}(\eg,R) / \rho(R)^2$ represents
the particle physics, where $q_{\rmn{sm},i}(\eg,R)$ is defined by
Eq.~(\ref{eq:q_sm}). The different density profiles have some impact
on the luminosity from annihilating DM, although the details of the
density profile can be neglected compared to the dominating boost from
substructures (assuming DM to be cold). In Fig.~\ref{fig:radial_lum}
we compare the radial dependence of the accumulative luminosity from
different smooth cluster density profiles to the boosted luminosity
due to substructure for different mass scales. We recalculate the
overdensity, $\rhos$, for the shallower density profile with
$\beta=0.6$ and rescale the concentration parameter in
Eq.~(\ref{eq:cfit}) with $300/160$ \cite{2011ApJ...728L..39N} to
account for the more centralized scale radius in cluster with a
shallow inner slope. The emission of this profile with $\beta=0.6$ is
about 30\% larger within $\rvir$ compared to a cuspy NFW
($\beta=1.0$).  This difference is built up within $0.1\rvir$
(i.e. close to $r_\s$). Hence the slope of the central part of a
cluster has little influence for the DM luminosity within $\rvir$ as
long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is negligible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. These boost factors are only realized for direct
annihilation emission (continuum emission or final state radiation) or
IC scattering off homogeneous seed photon fields (CMB). For IC
scattering of SD photons, the overlap of final state leptons and SD
photons is smaller which causes the substructure boost over the smooth
emission to be reduced by roughly two orders of magnitudes. It should
also be noted that the substructure boosted fluxes from clusters are
much more extended than for dwarf galaxies, hence more difficult to
detect with Cherenkov telescopes. In fact, it was shown in
Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that inside $0.1$~deg,
where the sensitivity of Cherenkov telescopes is maximized, the
expected DM flux including substructures from the brightest dwarf
galaxy is about an order magnitude higher than the brightest
cluster. However, that work assumed a substructure boost of about 50
in the most massive cluster halos which is more than an order
magnitude smaller than what we use in this work, hence in projection
we expect this difference to be a factor few smaller.

@@

In this section, we have found that the LP DM models overproduces both
the spatially extended differential and integrated gamma-ray flux
upper limits set by the \Fermi-LAT 18 month data for several
clusters. We can use these upper limits to constrain the boosts due to
SFE and substructures. Furthermore, as more sensitive experiments and
better upper limits emerge, we can start ruling out models that give
rise to boost factors that are manifested in the gamma-ray flux. It is
especially interesting to constrain the SFE, where a boost of the
particle physics cross section of $\sim$300 is required to explain the
excess of electrons and positrons observed in the vicinity of Earth by
PAMELA/\Fermi/H.E.S.S. In Fig.~\ref{fig:boost_const} we use the
estimated flux within $\rvir$ of four bright and maximal constraining
clusters (Fornax, M49, NGC4636, and Coma, where we have excluded Virgo
due to its large angular extent) to limit the SFE as well as the
minimum mass of substructures where
$M_\rmn{lim}\propto\mvir^{0.226}$. We find that M49 and the Fornax
cluster are the most constraining clusters, although the current upper
limits are not good enough to rule out the boost from substructures or
SFE. Assuming that the PAMELA positron excess is due to SFE DM, then
in order not to overproduce the differential upper limits set by
\Fermi-LAT in the energy range\footnote{We average the LP DM flux in
  the energy range $1-10$~GeV before we compare it to the upper
  limits.}  $1-10$~GeV, we can constrain the smallest mass of
substructures to $M_\rmn{lim}\gtrsim 10^4\,\msun$. Instead, assuming
the standard mass of the smallest substructures of $M_\rmn{lim}\approx
10^{-6}\,\msun$, we can constrain the saturated boost from SFE to
about \C{5 in M49 and Fornax}, which corresponds to a maximum SFE in the
MW of about 3. Hence, we conclude that without substantial improvement
in the modelling of these extended sources, \Fermi-LAT will not be
able to rule out LP models in their current form using clusters.

>>>>

\begin{abstract}
  We study the possibility for detecting gamma-ray emission from galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay, 2) different
  representative benchmark models of supersymmetric dark matter (DM), and 3)
  leptophilic models of DM annihilation that include a Sommerfeld enhancement
  (SFE). To model DM annihilation, we consider continuum emission as a result of
  the hadronization of annihilating neutralinos, internal bremsstrahlung, and
  inverse Compton emission from the cosmic microwave background as well as from
  a realistic spatial and spectral distribution of dust and stellar light. Among
  all clusters/groups, we predict Virgo, Fornax and M49 to be the brightest DM
  sources and find a particularly low CR-induced background for Fornax.  For a
  minimum substructure mass given by the DM free-streaming scale, cluster halos
  maximize the substructure boost for which we find a factor of $\gtrsim 1000$.
  Since the annihilation flux of substructures is mostly contributed by the
  regions around the virial radius, the resulting surface brightness profiles
  are almost flat. This makes it very challenging to detect this flux with
  imaging atmospheric Cherenkov telescopes since their sensitivity drops
  linearly with radius and they typically have $5-10$ linear resolution elements
  across a cluster; the gamma-ray satellite \Fermi is better suited.  Assuming
  cold dark matter with a substructure mass distribution down to an Earth mass
  and using \Fermi upper limits, we rule out the leptophilic models in their
  present form in 28 clusters, and limit the boost from SFE in M49 and Fornax to
  be $\lesssim 5$. This corresponds to a limit on SFE in the Milky Way of
  $\lesssim 3$. Alternatively, if the SFE is realized in Nature, this
  would imply a limiting substructure mass of $\mlim > 10^4\,\msun$---a
  challenge for structure formation. Without improved modelling of extended
  sources or stacked likelihood analyses, \Fermi will face difficulties in
  constraining our selection of DM benchmark models in the near future. The
  \Fermi upper limits are, however, closing in on our predictions for the CR
  flux using an analytic model based on cosmological hydrodynamical cluster
  simulations. We limit the CR-to-thermal pressure in nearby bright galaxy
  clusters of the \Fermi sample to $\lesssim 10\%$ and in Norma and Coma to
  $\lesssim 3\%$.  Thus we will soon start to constrain the underlying CR
  physics such as shock acceleration efficiencies or CR transport properties.
\end{abstract}

@@

Among the astrophysical systems which will be very interesting to detect, and
study, with gamma-ray detectors (\Fermi, H.E.S.S., MAGIC, VERITAS, and
eventually large detectors like CTA)\footnote{{High Energy Stereoscopic System,
    Major Atmospheric Gamma Imaging Cerenkov Telescope, Very Energetic Radiation
    Imaging Telescope Array System, Cerenkov Telescope Array.}} belong galaxy
clusters. The most promising directions in which to search for a gamma-ray
annihilation signal (from the annihilation process itself, and also the
accompanying bremsstrahlung and inverse Compton (IC) components coming from
charged particles produced in the annihilations) are basically three:

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. \C{However, if the boost from Sommerfeld
  enhancement (SFE) is small then the gamma-ray emission induced by
  CRs is dominating most clusters with the exception of a few
  clusters; most notably Fornax that shows the largest ratio of DM
  annihilation-to-CR flux. In particular, we identify Fornax and Virgo
  as prime targets for DM observations,} Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.}  For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ $^{(1)}$ & $\B_\rmn{sfe} $$^{(2)}$ &  $\B_\rmn{sub} $$^{(3)}$ \\
         & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ 
         & & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $8.2\times10^{-8}$  & $7.6\times10^{-11}$ & $4.1\times10^{-9}$  
 & \,\,\,\,\,         & $1.2\times10^{-11}$ & $2.8\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $12.9$ & $530/65$  & $1290$ \\
 Perseus \,\,\,\,\,\, & $8.6\times10^{-8}$  & $7.8\times10^{-11}$ & $1.5\times10^{-8}$  
 & \,\,\,\,\,         & $1.2\times10^{-11}$ & $2.9\times10^{-13}$ & $5.5\times10^{-12}$ 
 & $8.6$ & $530/75$  & $1190$ \\
 Virgo                & $1.6\times10^{-6}$  & $1.5\times10^{-9}$ & $1.5\times10^{-8}$  
 & \,\,\,\,\,         & $2.3\times10^{-10}$ & $5.4\times10^{-12}$ & $5.7\times10^{-12}$ 
 & $6.9$  & $530/80$  & $1120$  \\
 Fornax               & $3.5\times10^{-7}$  & $3.2\times10^{-10}$ & $3.1\times10^{-10}$ 
 & \,\,\,\,\,         & $5.1\times10^{-11}$ & $1.2\times10^{-12}$ & $1.1\times10^{-13}$ 
 & $2.4$  & $530/110$  & $890$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The mass of Fornax, Coma, and Perseus are derived from \cite{2007A&A...466..805C}, 
  while the mass of Virgo is derived from \cite{1984ApJ...281...31T}.\\
  (2) The boost due to Sommerfeld enhancement. The first value shows the saturated 
  Sommerfeld boost realized when substructures are present, the latter is the Sommerfeld 
  boost without substructures (see Section II A 1).\\
  (3) \C{The boost due to substructures relative a the smooth DM
  distribution.} We integrate the emission in a cylinder with a radial
  extent of $2.5 R_{200}$ along the line-of-sight and an angular size
  corresponding to $\rvir$  (both measured from the cluster center).\\
  (4) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  \Fermi (see the following sections for detail).\\
  (5) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (6) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). \C{Furthermore,
since most of the gamma-ray flux is expected to come from dense
subhalos within clusters which have 
a velocity dispersion}  close to the velocity limit
$v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross-section, but here we have additionally
  included the order unity constraint that ensures that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess at
Earth observed by the PAMELA and \Fermi satellites,
\begin{equation}
%\B_\sfe(\sigv) = 0.266\, \frac{c}{v_\rmn{sat}}\, \frac{v_\rmn{MW}}{\sigv} = 
%121 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.
\C{\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} = 
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.}
\label{eq:B_sfe}
\end{equation}
\C{Here the boost factor needed to explain the PAMELA/\Fermi excess of
  electrons and positrons $\B_\rmn{sfe,MW}=300$, the velocity
  dispersion for the MW $v_\rmn{MW} \approx 220\,\rmn{km\,s}^{-1}$, and the
  velocity dispersion $\sigv$ for clusters is derived in
  \cite{2005RvMP...77..207V}.} Clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution dissipationless DM simulations of MW type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center suffer from dynamical friction
and tidal effects depleting their central number densities. Since the
DM annihilation rate depends on the density squared, the resulting
flux from substructures is boosted compared to the smooth density
distribution.  While there is still a discrepancy in the literature of
the exact value of the predicted boost factor of the DM annihilation
luminosity from substructures in DM halos, this inconsistency starts
to become resolved, apparently converging toward predictions at the
high end \protect \cite{2008MNRAS.391.1685S, 2008Natur.456...73S,
  2010ApJ...718..899A}. \C{Following recent high-resolution
  simulations of Galaxy-sized halos, we adopt a boost factor due to
  substructures of 230 for such a halo \cite{2008MNRAS.391.1685S}
  (which needs to be scaled to cluster halos as we will discuss
  below).}

The initial suggestions of a small {\em total} boost ($<R_{200}$) of
order unity by the Via Lactea simulations \cite{2008JPhCS.125a2008K}
made assumptions of computing the boost assuming that the substructure
luminosity follows that of the smooth DM distribution. Physically,
this would imply substructure to follow the smooth DM distribution and
a radially independent concentration parameter of substructures. Both
assumptions are in conflict with tidal mass loss of satellites which
is at work in simulations and cause the radial number density to be
anti-biased with respect to the host's mass density profile yielding a
significantly flatter subhalo distribution compared to the smooth DM
distribution \cite{2010ApJ...718..899A}.  This increases the
substructure boost preferentially in the outer parts of DM halos with
a factor that ranges between 20 and 1000 (for dwarfs and galaxy
clusters), hence depending sensitively on the host halo mass (see
Figure~\ref{fig:radial_lum}). \C{Recent {\it Phoenix} simulations by
  Gao et al. \cite{2011arXiv1107.1916G} confirm the large enhancement
  due to substructures in clusters. In fact they find a boost of 1125
  for a Coma like cluster which agrees within 10\% with our estimated
  boost (see e.g. Table~\ref{tab:flux_tab}).}

@@

We now derive the squared density profile for the substructures
using
\begin{eqnarray}
\rho_\rmn{sub}^2(r) &=& \frac{\dd L_\rmn{sub}}{\dd V} \frac{1}{A_{\rmn{dm}}}\,,\label{eq:rho_sub}\\
L_{200\sm} &=&  A_{\rmn{dm}} \int_{\rvir} \dd r' 4\pi r'^2 \rho^2(r')\,,\label{eq:Lsm}\\
 \frac{\dd L_\rmn{sub}}{\dd V} &=& a_0\,a_1\,C(\mvir)\,x^{g(x)}\,
\left[\frac{1+a_2\log(x)}{4\pi\,r^3}\right]\,,\nonumber\\
\\
x &=& r/\rvir\,,\quad g(x) = a_2+a_1\,x^{a_2}\,.\label{eq:xvir}
\end{eqnarray} 
Here, $A_{\rmn{dm}} = q_{\rmn{sm},i}(\eg,R) / \rho(R)^2$ represents
the particle physics factor, where $q_{\rmn{sm},i}(\eg,R)$ is defined by
Eq.~(\ref{eq:q_sm}). The different density profiles have some impact
on the luminosity from annihilating DM, although the details of the
density profile can be neglected compared to the dominating boost from
substructures (assuming DM to be cold). In Fig.~\ref{fig:radial_lum}
we compare the radial dependence of the accumulative luminosity from
different smooth cluster density profiles to the boosted luminosity
due to substructure for different mass scales. We recalculate the
overdensity, $\rhos$, for the shallower density profile with
$\beta=0.6$ and rescale the concentration parameter in
Eq.~(\ref{eq:cfit}) with $300/160$ \cite{2011ApJ...728L..39N} to
account for the more centralized scale radius in cluster with a
shallow inner slope. The emission of this profile with $\beta=0.6$ is
about 30\% larger within $\rvir$ compared to a cuspy NFW
($\beta=1.0$).  This difference is built up within $0.1\rvir$
(i.e. close to $r_\s$). Hence the slope of the central part of a
cluster has little influence for the DM luminosity within $\rvir$ as
long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is negligible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. These boost factors are only realized for direct
annihilation emission (continuum emission or final state radiation) or
IC scattering off homogeneous seed photon fields (CMB). For IC
scattering of SD photons, the overlap of final state leptons and SD
photons is smaller which causes the substructure boost over the smooth
emission to be reduced by roughly two orders of magnitudes. It should
also be noted that the substructure boosted fluxes from clusters are
much more extended than for dwarf galaxies, hence more difficult to
detect with Cherenkov telescopes. In fact, it was shown in
Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that inside $0.1$~deg,
where the sensitivity of Cherenkov telescopes is maximized, the
expected DM flux including substructures from the brightest dwarf
galaxy is about an order magnitude higher than the brightest
cluster. However, that work assumed a substructure boost of about 50
in the most massive cluster halos which is more than an order
magnitude smaller than what we use in this work, hence in projection
we expect this difference to be a factor few smaller.

@@

In this section, we have found that the LP DM models overproduces both
the spatially extended differential and integrated gamma-ray flux
upper limits set by the \Fermi-LAT 18 month data for several
clusters. We can use these upper limits to constrain the boosts due to
SFE and substructures. Furthermore, as more sensitive experiments and
better upper limits emerge, we can start ruling out models that give
rise to boost factors that are manifested in the gamma-ray flux. It is
especially interesting to constrain the SFE, where a boost of the
particle physics cross section of $\sim$300 is required to explain the
excess of electrons and positrons observed in the vicinity of Earth by
PAMELA/\Fermi/H.E.S.S. In Fig.~\ref{fig:boost_const} we use the
estimated flux within $\rvir$ of four bright and maximal constraining
clusters (Fornax, M49, NGC4636, and Coma, where we have excluded Virgo
due to its large angular extent) to limit the SFE as well as the
minimum mass of substructures where
$M_\rmn{lim}\propto\mvir^{0.226}$. We find that M49 and the Fornax
cluster are the most constraining clusters, although the current upper
limits are not good enough to rule out the boost from substructures or
SFE. Assuming that the PAMELA positron excess is due to SFE DM, then
in order not to overproduce the differential upper limits set by
\Fermi-LAT in the energy range\footnote{We average the LP DM flux in
  the energy range $1-10$~GeV before we compare it to the upper
  limits.}  $1-10$~GeV, we can constrain the smallest mass of
substructures to $M_\rmn{lim}\gtrsim 10^4\,\msun$. Instead, assuming
the standard mass of the smallest substructures of $M_\rmn{lim}\approx
10^{-6}\,\msun$, we can constrain the saturated boost from SFE to
\C{$\lesssim 5$ in M49 and Fornax}, which corresponds to a maximum SFE
in the MW of $\lesssim3$. Hence, we conclude that without substantial
improvement in the modelling of these extended sources, \Fermi-LAT
will not be able to rule out LP models in their current form using
clusters but can put impressive constraints on them.
